---
title: "Exploratory Analysis of Suicide Data"
author: "Zikai Lin"
date: "11/8/2020"
output:
  html_document:
    number_sections: true
    toc: true
    code_folding: "hide"
    theme: readable
    highlight: haddock
---


# **Import & Tidy**

The majority of the data used in this analysis was obtained from the World Health Organisation.

***Data Cleaning Notes***

* 7 countries removed (<= 3 years of data total)
* 2016 data was removed (few countries had any, those that did often had data missing)
* HDI was removed due to 2/3 missing data
* Generation variable has problems, detailed in 2.11
* Continent was added to the dataset using the `countrycode` package
* Africa has very few countries providing suicide data

```{r message = FALSE, warning = FALSE, echo=FALSE}
library(tidyverse) # general
library(ggalt) # dumbbell plots
library(countrycode) # continent
library(rworldmap) # quick country-level heat maps
library(gridExtra) # plots
library(broom) # significant trends within countries

theme_set(theme_light())


# 1) Import & data cleaning

data <- read_csv("./master.csv") 


# data <- data %>% 
#   select(-c(`HDI for year`, `suicides/100k pop`)) %>%
#   rename(gdp_for_year = `gdp_for_year ($)`, 
#          gdp_per_capita = `gdp_per_capita ($)`, 
#          country_year = `country-year`) %>%
#   as.data.frame()
# 


# 2) OTHER ISSUES

# a) this SHOULD give 12 rows for every county-year combination (6 age bands * 2 genders):

# data %>% 
#   group_by(country_year) %>%
#   count() %>%
#   filter(n != 12) # note: there appears to be an issue with 2016 data
# not only are there few countries with data, but those that do have data are incomplete

# data <- data %>%
#   filter(year != 2016) %>% # I therefore exclude 2016 data
#   select(-country_year)


# b) excluding countries with <= 3 years of data:

# minimum_years <- data %>%
#   group_by(country) %>%
#   summarize(rows = n(), 
#             years = rows / 12) %>%
#   arrange(years)
# 
# data <- data %>%
#   filter(!(country %in% head(minimum_years$country, 7)))


# no other major data issues found yet



# # 3) TIDYING DATAFRAME
# data$age <- gsub(" years", "", data$age)
# data$sex <- ifelse(data$sex == "male", "Male", "Female")
# 
# 
# # getting continent data:
# data$continent <- countrycode(sourcevar = data[, "country"],
#                               origin = "country.name",
#                               destination = "continent")
# 
# # Nominal factors
# data_nominal <- c('country', 'sex', 'continent')
# data[data_nominal] <- lapply(data[data_nominal], function(x){factor(x)})
# 
# 
# # Making age ordinal
# data$age <- factor(data$age, 
#                    ordered = T, 
#                    levels = c("5-14",
#                               "15-24", 
#                               "25-34", 
#                               "35-54", 
#                               "55-74", 
#                               "75+"))
# 
# # Making generation ordinal
# data$generation <- factor(data$generation, 
#                    ordered = T, 
#                    levels = c("G.I. Generation", 
#                               "Silent",
#                               "Boomers", 
#                               "Generation X", 
#                               "Millenials", 
#                               "Generation Z"))
# 
# data <- as_tibble(data)

data = read.csv("./main_data.csv", as.is = T)


# the global rate over the time period will be useful:

global_average <- (sum(as.numeric(data$suicides_no)) / sum(as.numeric(data$population))) * 100000

```


# Global Trend

Here we want to investigate the global trend of the suicide data. 

```{r, message=F}
data %>%
  group_by(year) %>%
  summarize(population = sum(population), 
            suicides = sum(suicides_no), 
            suicides_per_100k = (suicides / population) * 100000) %>%
  ggplot(aes(x = year, y = suicides_per_100k)) + 
  geom_line(col = "deepskyblue3", size = 1) + 
  geom_point(col = "deepskyblue3", size = 2) + 
  geom_hline(yintercept = global_average, linetype = 3, color = "grey35", size = 1) +
  geom_vline(xintercept = 1995, color = "red", linetype = 4, size = 1)+
  geom_rect(aes(xmin=-Inf, xmax=1995, ymin=-Inf, ymax=Inf), alpha=0.003, color = "grey35")+
  labs(title = "Global Suicides (per 100k)",
       subtitle = "Trend over time, 1985 - 2015.",
       x = "Year", 
       y = "Suicides per 100k") + 
  scale_x_continuous(breaks = seq(1985, 2015, 2)) + 
  scale_y_continuous(breaks = seq(10, 20))
```

According to the plot, we be able to hypothesize that the trend of gobal suicide rate increase from 1985-1995 and decreased after 1995. 


# By continents
However, this is just a global trend, in order to investigate if the trends are different within differnt continents. We can plot the trend of suicide rate for different continents.


```{r, message=FALSE}
continent <- data %>%
  group_by(continent) %>%
  summarize(suicide_per_100k = (sum(as.numeric(suicides_no)) / sum(as.numeric(population))) * 100000) %>%
  arrange(suicide_per_100k)

continent$continent <- factor(continent$continent, ordered = T, levels = continent$continent)


continent_time <- data %>%
  group_by(year, continent) %>%
  summarize(suicide_per_100k = (sum(as.numeric(suicides_no)) / sum(as.numeric(population))) * 100000)

continent_time$continent <- factor(continent_time$continent, ordered = T, levels = continent$continent)

continent_time_plot <- ggplot(continent_time, aes(x = year, y = suicide_per_100k, col = factor(continent))) + 
  facet_grid(continent ~ ., scales = "free_y") + 
  geom_line() + 
  geom_point() + 
  geom_vline(xintercept = 1995, linetype=3)+
  labs(title = "Trends Over Time, by Continent", 
       x = "Year", 
       y = "Suicides per 100k", 
       color = "Continent") + 
  theme(legend.position = "none", title = element_text(size = 10)) + 
  scale_x_continuous(breaks = seq(1985, 2015, 5), minor_breaks = F)
continent_time_plot

continent_time_plot_one <- ggplot(continent_time, aes(x = year, y = suicide_per_100k, col = factor(continent))) + 
  geom_line() + 
  geom_point() + 
  geom_vline(xintercept = 1995, linetype=3)+
  labs(title = "Trends Over Time, by Continent", 
       x = "Year", 
       y = "Suicides per 100k", 
       color = "Continent") + 
  theme(title = element_text(size = 10)) + 
  scale_x_continuous(breaks = seq(1985, 2015, 5), minor_breaks = F)
continent_time_plot_one


```
It seems that 

* 1995 seems to be the turning point for Africa, Oceania and Europe
* However, for America, the suicide rate keep increasing.
* The suicide rate of Asia seems to be consistent over time 

Based on the above analysis, we might want to introduce the $$\mathrm{Continent}\times \mathrm{Time}_{ij}$$ interaction, because the for different continent, the trend of suicide rate seems to be different.



# Age difference


```{r, message=FALSE}

### with time
age_time_plot <- data %>%
  group_by(year, age) %>%
  summarize(suicide_per_100k = (sum(as.numeric(suicides_no)) / sum(as.numeric(population))) * 100000) %>%
  ggplot(aes(x = year, y = suicide_per_100k, col = age)) + 
  geom_line() + 
  geom_point() + 
  labs(title = "Trends Over Time, by Age", 
       x = "Year", 
       y = "Suicides per 100k", 
       color = "Age") + 
  scale_x_continuous(breaks = seq(1985, 2015, 5), minor_breaks = F)
age_time_plot

```

# Gender difference
```{r, message=FALSE}
data %>%
  group_by(year, sex) %>%
  summarize(suicide_per_100k = (sum(as.numeric(suicides_no)) / sum(as.numeric(population))) * 100000) %>%
  ggplot(aes(x = year, y = suicide_per_100k, col = sex)) + 
  geom_line() + 
  geom_point() + 
  labs(title = "Trends Over Time, by Sex", 
       x = "Year", 
       y = "Suicides per 100k", 
       color = "Sex") + 
  scale_x_continuous(breaks = seq(1985, 2015, 5), minor_breaks = F)

```

# GDP as a random effect

In [R-Exploratory](https://www.kaggle.com/lmorgan95/r-suicide-rates-in-depth-stats-insights?scriptVersionId=14410354), Morgan explored the potential correlation effect between suicide rate and GDP.


```{r, message=FALSE}
country_mean_gdp <- data %>%
  group_by(country, continent) %>%
  summarize(suicide_per_100k = (sum(as.numeric(suicides_no)) / sum(as.numeric(population))) * 100000, 
            gdp_per_capita = mean(gdp_per_capita))

model1 <- lm(suicide_per_100k ~ gdp_per_capita, data = country_mean_gdp)

gdp_suicide_no_outliers <- model1 %>%
  augment() %>%
  arrange(desc(.cooksd)) %>%
  filter(.cooksd < 4/nrow(.)) %>% # removes 5/93 countries
  inner_join(country_mean_gdp, by = c("suicide_per_100k", "gdp_per_capita")) %>%
  select(country, continent, gdp_per_capita, suicide_per_100k)

ggplot(gdp_suicide_no_outliers, aes(x = gdp_per_capita, y = suicide_per_100k, col = continent)) + 
  geom_point() + 
  geom_smooth(method = "lm", aes(group = 1)) + 
  scale_x_continuous(labels=scales::dollar_format(prefix="$"), breaks = seq(0, 70000, 10000)) + 
  labs(title = "Correlation between GDP (per capita) and Suicides per 100k", 
       subtitle = "Plot with high CooksD countries removed (5/93 total)",
       x = "GDP (per capita)", 
       y = "Suicides per 100k", 
       col = "Continent")
```

And conclude that the linear association between suicide rate and GDP is significant but only account for a small amount of variation. However, by simply looking at the scatter plot, we don't have strong evidence to assume that the trend is linear. So we might want to treat GDP as a random slope.

```{r}
ggplot(gdp_suicide_no_outliers, aes(x = gdp_per_capita, y = suicide_per_100k)) + 
  geom_point() + 
  geom_smooth(method = "loess", aes(group = 1)) + 
  scale_x_continuous(labels=scales::dollar_format(prefix="$"), breaks = seq(0, 70000, 10000)) + 
  labs(title = "Correlation between GDP (per capita) and Suicides per 100k", 
       subtitle = "Plot with high CooksD countries removed (5/93 total)",
       x = "GDP (per capita)", 
       y = "Suicides per 100k", 
       col = "Continent")
```


# Model 

Based on the exploratory analysis, I think I will propose the following longitudinal model

$$\begin{aligned}
E(Y_{ij}) = &\beta_0 + \beta_1 \mathrm{Continent}_i + \beta_2 \mathrm{Sex}_i +  \beta_3 \mathrm{Year}_{j}+\beta_4 (\mathrm{Year}_{ij} - 1995)_+ +\beta_5 \mathrm{Continent}_i\times \mathrm{Year}_{j} + \beta_6 \mathrm{Sex}_i\times \mathrm{Year}_{j} \\
&+ \beta_7 \mathrm{Sex}_i\times (\mathrm{Year}_{ij} - 1995)_+ + \beta_8 \mathrm{Continent}_i\times (\mathrm{Year}_{ij} - 1995)_+ + b_{1i} + b_{2i} \mathrm{Year}_{j} + b_{3i} (\mathrm{Year}_{ij} - 1995)_+ + b_{4i} \mathrm{GDP}_{ij} + \epsilon_{ij}
\end{aligned}
$$

# Remaining questions to address

* Model interpretation
* The specific knot in linear spline, 1995, 1996 or 2000?
* Model comparison
* Model diagnostic
* Model fitting method (REML, ML, GEE?)